---
title: "Bathymetry Mapping with R"
author: "Marko Lipka"
output: 
  html_document: 
    fig_height: 10
    fig_width: 12
    theme: spacelab
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

### Get bathymetry data

The *marmap* package provides world wide high resolution bathymetry data.

```{r getBathymetryData}
library("marmap")
library("ggplot2")
get.bathymetry <- function(lon1=4, lon2=30, lat1=53, lat2=67, bathy.breaks = 8){
  bathymetry <- getNOAA.bathy(lon1=lon1, lon2=lon2, lat1=lat1, lat2=lat2, resolution=1, keep=TRUE) # keep = TRUE saves downloaded data as csv-File
  fortyfied.bathy <- fortify(bathymetry) # make a df out of bathy so ggplot can fully use the data
  fortyfied.bathy <- subset(fortyfied.bathy, z <= 0) # limit to values below sea surface
  fortyfied.bathy$z <- -fortyfied.bathy$z # make depths positive values
  names(fortyfied.bathy) <- c("Longitude", "Latitude", "Depth_m")
  fortyfied.bathy$Depthsteps_m <- cut(fortyfied.bathy$Depth_m, breaks = bathy.breaks, dig.lab = 4) # generate depth intervals for contour plot
  ## 'cut()' returns intervals in unpleasant format. Thus, cumbersome renaming for nice legend:
  fortyfied.bathy <- subset(fortyfied.bathy, !is.na(fortyfied.bathy$Depthsteps_m))
    levels(fortyfied.bathy$Depthsteps_m) <- sub(",", " - ", levels(fortyfied.bathy$Depthsteps_m))
    levels(fortyfied.bathy$Depthsteps_m) <- sub("\\(", "", levels(fortyfied.bathy$Depthsteps_m))
    levels(fortyfied.bathy$Depthsteps_m) <- sub("\\]", "", levels(fortyfied.bathy$Depthsteps_m))
    levels(fortyfied.bathy$Depthsteps_m) <- sub("(.*) - Inf", ">\\1", levels(fortyfied.bathy$Depthsteps_m))
  return(fortyfied.bathy)
}
```

```{r}
baltic.bathy   <- get.bathymetry(lon1 = 4, lon2 = 30, 
                                 lat1 = 53, lat2 = 67,
                                 bathy.breaks = c(seq(0, 50, length.out = 6),
                                                  seq(100, 300, length.out = 3),
                                                  +Inf))

ggplot(baltic.bathy) + 
    geom_tile(aes(x=Longitude, y=Latitude, fill=Depthsteps_m)) +
    scale_fill_brewer(palette = "Blues", name = "Water depth [m]") 
```


### Load world coastlines

The *mapdata* package allows to download world coastline polygon data. Unfortunately these are missing some 'small' islands (like Hiddensee).
Haven't found a workaround for that yet...

```{r getCoastlines}
library("mapdata")
baltic.coastlines <- map_data('worldHires', xlim = c(4, 29), ylim = c(50, 66))

ggplot(baltic.coastlines) +
  geom_polygon(aes(x=long, y=lat, group=group), colour = "white")
```


### Plot some maps

```{r plotFunction}
 

plot.map <- function(lon.min = 4, lon.max = 30,
                     lat.min = 53, lat.max = 67,
                     bathy.breaks = c(seq(0, 50, length.out = 6),
                                                  seq(100, 300, length.out = 3),
                                                  +Inf),
                     land.colour = NA, border.colour = "darkgrey"){
    bathy <- get.bathymetry(lon1 = lon.min, lon2 = lon.max, 
                                 lat1 = lat.min, lat2 = lat.max,
                                 bathy.breaks = bathy.breaks)
    coastlines <- map_data('worldHires', xlim = c(lon.min, lon.max), ylim = c(lat.min, lat.max))
    
    ggplot() +
    coord_quickmap(#projection= "azequalarea",
      xlim=c(lon.min, lon.max), ylim=c(lat.min, lat.max)) +
    geom_tile(data = bathy, aes(x=Longitude, y=Latitude, fill=Depthsteps_m)) +
    scale_fill_brewer(palette = "Blues", name = "Water depth [m]") +
    geom_polygon(data=coastlines, aes(x=long, y=lat, group=group),
                 fill=land.colour, colour = border.colour, lwd=.5) +
        scale_y_continuous(expand = c(0,0)) + scale_x_continuous(expand = c(0,0)) + # so that the whole plot area is filled with bathy and coastlines.
    theme_minimal()
}
```

#### The whole Baltic Sea:

```{r Map_BalticSea}
plot.map()
```

#### The southern part only:

... with the bathymetry depth range of the whole Baltic Sea ...

```{r Map_SouthernBalticSea}
plot.map(lon.min = 4, lon.max = 30,
         lat.min = 53, lat.max = 67) +
    coord_quickmap(xlim = c(10, 20), ylim = c(54, 58))
```

... or adapted to plot region.

```{r Map_SouthernBalticSea_ownDepRange}
plot.map(lon.min = 10, lon.max = 20, lat.min = 54, lat.max = 58)
```

#### Should also work for any other region in the world

```{r Map_randomRegion}
long <- 110
lat <- 5

plot.map(lon.min = long, lon.max = long +40, lat.min = lat, lat.max = lat + 15, 
         bathy.breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000, 7000, +Inf),
         land.colour = "grey20", border.colour = "pink")
```

Viola! Have fun ...

## Known issues / \#TODO:

* map_data even with *map = 'worldHires'* does not include all islands:
  + is there another source for high resolution coastlines of the world?
  + is the coastline necessary ?
* would be nice to have some options for the colour scale of bathymetry
    + continuous vs. discrete
    + colour gradient selection
    + more convenient labeling / interval selection