---
title: "Bathymetry mapping with ggplot in R"
author: "Marko Lipka"
output: 
  html_document: 
    fig_height: 10
    fig_width: 12
    theme: spacelab
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

### Get bathymetry data

The *marmap* package provides world wide high resolution bathymetry data.

```{r getBathymetryData}
source("ggPlotBathymetry.R")
```

```{r}
baltic.bathy   <- get.bathymetry(lon1 = 4, lon2 = 30, 
                                 lat1 = 53, lat2 = 67,
                                 bathy.breaks = c(seq(0, 50, length.out = 6),
                                                  seq(100, 300, length.out = 3),
                                                  +Inf))

ggplot(baltic.bathy) + 
    geom_tile(aes(x=Longitude, y=Latitude, fill=Depthsteps_m)) +
    scale_fill_brewer(palette = "Blues", name = "Water depth [m]") 
```


### Load world coastlines

The *mapdata* package allows to download world coastline polygon data. Unfortunately these are missing some 'small' islands (like Hiddensee).
Haven't found a workaround for that yet...

```{r getCoastlines}
baltic.coastlines <- map_data('worldHires', xlim = c(4, 29), ylim = c(50, 66))

ggplot(baltic.coastlines) +
  geom_polygon(aes(x=long, y=lat, group=group), colour = "red", fill = NA)
```


### Plot some maps

... combine plot of bathymetry and coastlines in one function *plot.bathymetry()*

#### The whole Baltic Sea:

```{r Map_BalticSea}
plot.bathymetry()
```

#### The southern part only:

... with the bathymetry depth range of the whole Baltic Sea ...

```{r Map_SouthernBalticSea}
plot.bathymetry(lon.min = 4, lon.max = 30,
         lat.min = 53, lat.max = 67) +
    coord_quickmap(xlim = c(10, 20), ylim = c(54, 58))
```

... or adapted to plot region.

```{r Map_SouthernBalticSea_ownDepRange}
plot.bathymetry(lon.min = 10, lon.max = 15, lat.min = 54, lat.max = 56)
```

#### Should also work for any other region in the world

```{r Map_randomRegion}
long <- 110
lat <- 5

plot.bathymetry(lon.min = long, lon.max = long +40, lat.min = lat, lat.max = lat + 15, 
         bathy.breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000, 7000, +Inf),
         land.colour = "grey20", border.colour = "pink")
```

Viola! Have fun ...

## Known issues / \#TODO:

* map_data even with *map = 'worldHires'* does not include all islands:
    + is there another source for high resolution coastlines of the world?
    + is the coastline necessary ?
* would be nice to have some options for the colour scale of bathymetry
    + continuous vs. discrete
    + colour gradient selection
    + more convenient labeling / interval selection