---
title: "Bathymetry mapping with ggplot in R"
author: "Marko Lipka"
output: 
  pdf_document: 
    dev: png
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)
```

### Intro


```{r getBathymetryData}
source("ggPlotBathymetry.R")
```

The  provides world wide high resolution bathymetry data.

The  allows to download world coastline polygon data. 

*plot.bathymetry()* fetches bathymetry data (*marmap* package) and coastlines (*mapdata* package) from the desired region and returns a ggplot that can be subsequently extended by further ggplot-layers.


## Arguments

argument | description 
---------|------------------------------------------------------------------------------
lon.min, lon.max, lat.min, lat.max | Limits of the region to be plotted in decimal degrees
bathy.breaks | either a numeric vector up to 9 unique cut points or a single number (up to 8) giving the number of intervals for the bathymetry plot
land.colour, border.colour | valid name or number specifying a color (defaults are *NA* for land.color and "black" for border.color)

\newpage

### Examples

#### The Baltic Sea:
... with manually defined bathymetry break points.
```{r Map_BalticSea, fig.width=10, fig.height=8}
plot.bathymetry(lon.min = 4, lon.max = 30,
                lat.min = 53, lat.max = 67,
                bathy.breaks = c(seq(0, 50, length.out = 6),
                                 seq(100, 300, length.out = 3),
                                 +Inf))
```

\newpage

#### The southern Baltic Sea region:

... with coloured dry land and additional plot layers.

```{r Map_SouthernBalticSea_ownDepRange, fig.width=10, fig.height=5}
plot.bathymetry(lon.min = 10, lon.max = 15, lat.min = 54, lat.max = 56, 
                bathy.breaks = seq(0, 100, 10),
                land.colour = "tan2") +
    geom_line(data = data.frame(lat = c(54.2, 54.5, 54.8, 55, 55.5),
                                 lon = c(11.5, 12, 13.1, 14.2, 14.8)),
               mapping = aes(x = lon, y = lat),
               color = "black", size = 3, linetype = "twodash") +
    annotate(geom = "text", x = 13.7, y = 55, angle = 20,
             label = "Arkona Basin")
```

\newpage

To get a plot of the southern Baltic Sea while maintaining the range of the legend for the whole Baltic Sea (see above), plot range of the 'whole Baltic Sea plot' can be limited via 'coord_quickmap()':

```{r Map_SouthernBalticSea, fig.width=10, fig.height=5}
plot.bathymetry(lon.min = 4, lon.max = 30,
         lat.min = 53, lat.max = 67,
         bathy.breaks = c(seq(0, 50, length.out = 6),
                                 seq(100, 300, length.out = 3),
                                 +Inf),
         land.colour = "black") +
    coord_quickmap(xlim = c(10, 15), ylim = c(54, 56))
```

\newpage

#### Works for any region in the world

```{r Map_randomRegion, fig.width=10, fig.height=3}
long <- 110
lat <- 5

plot.bathymetry(lon.min = long, lon.max = long +40, lat.min = lat, lat.max = lat + 15, 
         bathy.breaks = c(0, 1000, 2000, 3000, 4000, 5000, 6000, 7000, +Inf),
         land.colour = "grey20", border.colour = NA)
```


## Known issues / \#TODO:

* map_data even with *map = 'worldHires'* does not include all islands:
    + is there another source for high resolution coastlines of the world?
    + is the coastline necessary ?
* would be nice to have some options for the colour scale of bathymetry
    + continuous vs. discrete
    + colour gradient selection
    + more convenient labeling / interval selection
* option for (labeled) contour lines